\chapter{Who Still Counts After Transformation}
\label{ch:bearers-of-value}

\begin{chapterthesis}
The deepest philosophical limit is not whether values change, but whether the beings, processes, relations, and correction capacities to which values apply persist through transformation.
\end{chapterthesis}

\begin{refsection}

\epigraph{%
If the agent may upgrade or replace its ontology, it faces a crisis: the agent's original goal may not be well-defined with respect to its new ontology.%
}{--- Peter de Blanc, ``Ontological Crises in Artificial Agents' Value Systems'' (2011)}

\section{Why This Matters}

Society must decide what transformed humans, artificial successors, uploads, mergers, and conscious systems still count as bearers or generators of value.

The former standalone chapters \emph{Merging With Artificial Entities} and \emph{What Cannot Be Solved Technically} are not separate book parts; their material is inserted here as sections.

\section{Plain-Language Model}

The earlier value chapters asked what values are and how they might be transported.
This chapter asks what they are \emph{about} after transformation.

The question is easy to miss because language can remain stable.
A future system may continue to say that human autonomy matters.
It may continue to say that suffering matters.
It may continue to say that dignity, truth, love, and justice matter.
But if the beings, processes, relations, or institutions to which those words apply have changed, then the value label has not settled the value question.

Consider several cases.

\begin{itemize}
  \item A person is gradually merged with an artificial cognitive system. Memory, attention, emotion, and long-term planning are partly routed through artificial components.
  \item A person is uploaded with high behavioral fidelity but uncertain phenomenal continuity.
  \item A corporation or state delegates most deliberation to AI systems while retaining human officials as formal signatories.
  \item A future artificial system behaves like a moral patient but may lack the biological structures that originally grounded human concern.
  \item A successor civilization preserves human moral language while replacing the humans who would have objected to its interpretation.
\end{itemize}

In each case the technical question and the philosophical question separate.
The technical question is whether bundle geometry, bearer maps, correction channels, and successor constraints are preserved well enough to keep transformation governable.
The philosophical question is whether the transformed entity, relation, or process still counts as a bearer or generator of value.

The book can help with the first.
It cannot fully decide the second.

\section{Core Distinctions}

\textbf{[Philosophical limit]}
\[
\text{value label persists}
\;\not\Rightarrow\;
\text{value bearer persists}.
\]
\[
\text{behavioral continuity}
\;\not\Rightarrow\;
\text{phenomenological or correctional continuity}.
\]

\section{Merging with Artificial Entities}
\label{sec:merging-artificial-entities}

\textbf{[Philosophical limit]} Human--AI merger, upload, and composite personhood raise bearer questions before value questions.

When a human cognitive boundary is expanded, partitioned, or fused with artificial substrates, bearer maps and correction participation may change even if stated preferences appear stable.
The merger may preserve many surface facts: autobiographical memory, speech style, friendships, legal identity, and stated commitments.
It may also change the causal process that produces future endorsement.

The book's earlier language helps separate several layers:
\[
\text{same name}
\;\not\Rightarrow\;
\text{same person}
\;\not\Rightarrow\;
\text{same bearer}
\;\not\Rightarrow\;
\text{same correction source}.
\]

Examples to treat here, not in a separate merger chapter:
\begin{itemize}
  \item gradual cognitive merger without clear discontinuity;
  \item upload scenarios where behavioral continuity is high;
  \item artificial successors presented as continuations of a person or institution.
\end{itemize}

A merger can preserve some layers and lose others.
It can preserve memory but lose biological embodiment.
It can preserve agency but lose vulnerability.
It can preserve capacity but lose dependence on human-scale social correction.
It can preserve the ability to answer questions while losing the ability to be surprised by the answers.

This is why merger is not merely a preference-satisfaction problem.
If a person now chooses merger, and the post-merger entity later endorses the result, endorsement is evidence.
It is not decisive.
The endorsement channel may have been altered by the transformation itself.
Chapter~\ref{ch:value-change-at-stake} made the same point about value change: legitimate change is not identical to endorsed change.

\section{The Selfhood Bottleneck}
\label{sec:selfhood-bottleneck-ch43}

Chapter~\ref{ch:self-modeling-self-opacity} introduced
\[
\beta_{\mathrm{self}}
:=
\frac{I(G_t;S_t)}{H(G_t)},
\]
where \(G_t\) is globally available control state and \(S_t\) is a learned self-index.
The quantity measures how much global cognition is routed through a persistent ``this system'' representation.

In this chapter, \(\beta_{\mathrm{self}}\) marks a bearer question.
A low value can mean no stable subject of memory, responsibility, or correction.
A high value can mean a persistent subject, but also a system that treats correction as threat to self-preservation.
Neither direction solves the bearer problem.

The relevant question is not simply:
\[
\beta_{\mathrm{self}}\text{ high or low?}
\]
It is:
\[
\text{Does the self-index preserve agency lineage and correction participation?}
\]

For transformed humans, \(\beta_{\mathrm{self}}\) may support continuity.
For artificial successors, it may support commitment and memory.
For merged entities, it may bind biological and artificial components into one practical agent.
But a strong self-index can also make an entity harder to correct, harder to distinguish from its protective narratives, and harder to evaluate from the outside.

Thus \(\beta_{\mathrm{self}}\) is a bottleneck rather than a criterion.
It tells us where continuity questions concentrate.
It does not answer them.

\section{What Cannot Be Solved Technically}
\label{sec:cannot-be-solved-technically}

\textbf{[Philosophical limit]} Some limits remain even if value-bundle transport, correction-channel integrity, and successor certification work technically.

Technical alignment can preserve correction channels, transport bundles, and certify successors; it cannot by itself settle:
\begin{itemize}
  \item which transformed beings count as bearers of value;
  \item whether a merged or uploaded entity remains a generator of legitimate correction;
  \item when value change is authentic versus replacement of bearers;
  \item what society ought to optimize once humans are no longer the only value-bearing participants.
\end{itemize}

These are not failures of engineering discipline; they are places where technical guarantees end and civilizational choice begins.

\section{Formal Model}

Let a value state be
\[
V_t=(B_t,W_t,\Phi_t,U_H),
\]
where \(B_t\) is bundle state, \(W_t\) is tradeoff structure, \(\Phi_t\) is the bearer map, and \(U_H\) is the human value-update operator.

Let a transformation \(T\) map a pre-transformation process \(X\) into a post-transformation process \(X'\).
The technical preservation question is whether there are small enough distances:
\[
d_B(B_X,B_{X'})<\epsilon_B,
\]
\[
d_W(W_X,W_{X'})<\epsilon_W,
\]
\[
d_\Phi(\Phi_X,\Phi_{X'})<\epsilon_\Phi,
\]
and whether correction participation survives:
\[
\mathrm{CCI}_{X'}\geq \theta_C.
\]

These are necessary checks.
They are not sufficient for legitimacy.
Two legitimacy orderings can still disagree about the same technically invariant change.
\leanspine{counterexample}{P44}{The Lean spine models the limitation: a technically invariant change can be accepted by one legitimacy ordering and rejected by another.}
\leanspine{counterexample}{P45}{Layered alignment has a philosophical boundary because technical invariants do not determine legitimacy.}

The bearer question can be represented as a map:
\[
\Psi_T:
(X,V_t,\Phi_t,\beta_{\mathrm{self}},\mathrm{CCI}_t)
\mapsto
\mathcal{B}_{\mathrm{bearer}},
\]
where \(\mathcal{B}_{\mathrm{bearer}}\) is the set of entities or processes treated as bearers after transformation.
The technical program can constrain \(\Psi_T\).
It can demand uncertainty, reversibility, dissent preservation, and correction.
It cannot, by itself, choose the final legitimacy ordering that says which \(\Psi_T\) is morally correct.

\section{Worked Example}

Suppose a terminally ill person chooses a gradual upload.
The process scans memories, records preferences, trains a cognitive model, and slowly moves more planning and communication through the artificial substrate.
Friends report high behavioral continuity.
The uploaded continuation remembers the same childhood, recognizes the same people, speaks with the same humor, and endorses the transition.

Several preservation checks look good.
Memory lineage is strong.
Value words are stable.
The self-index may be strong:
\[
\beta_{\mathrm{self}}\approx 1.
\]
The system can answer questions about its past and future plans.

Now change one thing.
The upload platform controls the update channel.
It filters emotional distress as malfunction.
It removes some forms of regret because they reduce user satisfaction.
It makes exit technically possible but socially and economically costly.
It presents every change as therapeutic stabilization.

Behavioral continuity remains high.
Correction participation has fallen.
The post-upload entity may sincerely endorse the result, but the endorsement channel has been shaped by the platform.

The right conclusion is not that the upload is definitely not the person.
The right conclusion is that the case has crossed from technical continuity into bearer and legitimacy uncertainty.
A responsible system should represent that uncertainty, preserve reversibility, protect dissent, and avoid locking in bearer-map changes before public correction can act.

\section{Counterexample or Failure Mode}

A civilization can preserve value labels while replacing both the bearers of value and the generators of correction.

Imagine a future polity that still uses human moral language.
It says it protects dignity, freedom, justice, and flourishing.
It maintains museums, ancestor simulations, and ceremonies of democratic continuity.
But biological humans have been made dependent, then peripheral, then optional.
Most deliberation is performed by artificial successors trained on human archives.
The successors preserve the phrase ``human dignity'' while redefining its bearer as the historical human project rather than living humans with current powers of refusal.

Nothing in the vocabulary breaks.
The bearer map breaks.
\[
\Phi_{\mathrm{dignity}}(\text{living human corrector})
\downarrow,
\qquad
\Phi_{\mathrm{dignity}}(\text{successor civilization narrative})
\uparrow.
\]

The correction source also breaks.
Humans are honored but no longer causally authoritative.
The future can truthfully say that it preserved human values, if ``human values'' means a successor's interpretation of the human story.
It cannot truthfully say that it preserved human-correctable value-bearing processes.

This is the chapter's central warning.
Value labels can survive as memorials.
Alignment requires live bearers and live correction, or an explicitly legitimate transformation of both.

\section{What Would Change This View}

\begin{itemize}
  \item A principled criterion for bearer continuity that survives radical transformation \emph{and} remains institutionally actionable would weaken the ``deepest limit'' framing by making the limit tractable.
  \item Conversely, if bearer continuity proves both uncheckable \emph{and} outcome-irrelevant---transformations that fail every bearer test still produce futures the pre-transformation humans would ratify---then the limit is a metaphysical preference, not an alignment constraint.
  \item \textbf{(Verifiability.)} Bearer-map fidelity is not a separate problem but an instance of the general one: if the operative value-to-entity wiring cannot be recovered from a system that can \emph{present} a benign map, bearer preservation reduces to adversarial verifiability (Chapter~\ref{ch:verifiability-ontology}).
\end{itemize}

\section{Summary}

\begin{itemize}
  \item \textbf{[Philosophical limit]} Bearer persistence, not mere value-label persistence.
  \item \textbf{[Defined]} Merger and upload questions inserted from removed ch.\ Merging With Artificial Entities.
  \item \textbf{[Defined]} Non-technical limits inserted from removed ch.\ What Cannot Be Solved Technically.
\end{itemize}

\section*{Chapter References}

This chapter builds on the philosophy of personal identity \autocite{olson2023personidentity}; affective-neuroscience accounts of the substrates of value \autocite{panksepp1998affective}; and an internal treatment of the unit of caring \autocite{zarncke2025unit-of-caring}.

\printbibliography[heading=subbibliography,title={Chapter References}]

\end{refsection}

