\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

Many lines of argument in this book grew out of work on operational agent boundaries and unsupervised agent discovery.
Several people shaped that foundation directly.

Jonas Hallgreen and I discussed agent foundations at length: taxonomies of agency, and different lenses on what an agent is, from crystalline structure to dynamical systems and beyond.
Those conversations sharpened the book's insistence that alignment must begin by finding the real optimizer, not by assuming it.

Chris Pang and Peter Kuhn have developed derivative work on agent discovery, especially the case in which one agent models another.
Talks with them, and with Jonas, pushed extensions of unsupervised agent discovery in several directions: agents as operators, the consciousness of agents, and tighter formalization.
This manuscript owes a substantial debt to that thread.

Brad Clark and the Foresight Institute Berlin team hosted presentations, talks, and workshops where these ideas were tested in public.
Foresight's support made sustained work on agent discovery and alignment possible.
The inquiry began at AE Studio, where I first pursued agent foundations in a product and research setting; I am grateful for the freedom to continue that line independently afterward.

Cameron Berg, Diogo de Lucena, and Alex McKenzie gave early reviews that improved clarity and claim strength when the framework was still forming.

SJ Beard's work on parameters of agency and the philosophical grounding of agency, and our discussions in a shared project, supplied inspiration that runs through the book's treatment of boundaries and bearers.

Broader communities kept the questions alive.
LessWrong discussions, the effective-altruism meetup group in Hamburg (especially Andreas Jessen and Ruslan), and the co-founders of Aintelope (Andre Kochanke, Hauke Rehfeld, Joel Pyykkö, Angie Normandale, Rasmus Herlo, and Roland Pihlakas) provided years of informal pressure-testing.
The PIBBSS team, participants in the Human Alignment Summer School in Prague, and colleagues in the AI Safety Group in Berlin offered further encouragement and critique.

Any remaining errors of fact, inference, or emphasis are mine alone.
