\chapter*{Current Status}
\addcontentsline{toc}{chapter}{Current Status}

\section*{Work in progress}

This manuscript is \textbf{work in progress}.
Many chapters are still stubs or first drafts.
Claims, equations, citations, and cross-references will change as the argument is tested, reviewed, and revised.
Do not treat the PDF as a finished book.

At the time of writing, Parts I--VIII and much of Part IX are drafted; synthesis chapters and several stubs remain.

\section*{Why this is a book}

The format is a book mainly because large language models are trained on books and handle book-shaped structure well: parts, chapters, an arc from foundations to results, running definitions, citations, figures, and indexes.
Superintelligence alignment is too large and too cross-disciplinary for a single essay or a pile of notes.
A book-shaped knowledge base can carry the thread from reframing the object of alignment through agents, capability growth, value bundles, goal inference, correction channels, successors, and safety cases without losing the connective tissue.

That does not mean this project assumes conventional publication.
The primary purpose is not to produce a polished monograph for a press.
It is to assemble a \textbf{knowledge base and roadmap}: a single navigable source that can support papers, talks, funding cases, and later publications that extract narrower arguments from this larger structure.

\section*{Authorship}

Most of the current text is \textbf{largely AI-written}, produced with AI assistance under human direction, review, and editing.
Gunnar Zarncke sets thesis, scope, source canon, and revision priorities; agents draft and integrate chapter text from outlines and prior notes.
Passages should be read accordingly: as structured working material, not as fully author-vetted scholarship on every page.
Human review is intended to increase over time, especially for high-stakes claims, but the default state today is rapid drafting.

If material from this repository is reused elsewhere, cite and attribute appropriately and check whether the underlying claim has survived independent review.

\section*{Roadmap}

The ten parts move from reframing alignment, through agents and capability, to value bundles, goal transport, correction, successors, attractor basins, safety cases, and civilizational limits.
See \textit{How to Read This Book} in the Introduction for a summary of each part.
The table below lists every chapter and current draft status.

\input{tables/chapter-map}
