\chapter*{Executive Overview}
\addcontentsline{toc}{chapter}{Executive Overview}

\epigraph{%
Only variety can destroy variety.%
}{--- W.\ Ross Ashby, \emph{An Introduction to Cybernetics} (1956)}

\section*{TL;DR}

\begin{itemize}
  \item Superintelligence alignment preserves \emph{human-correctable value-bearing processes} across capability growth, ontology shift, successor creation, and strategic multi-agent selection pressure---assuming civilization retains enough correction capacity to participate.
  \item The central theory tracks five linked preservation problems as systems grow, change ontology, and create successors:
  \begin{itemize}
    \item after transformation, do the same compressed value \emph{directions} still shape behaviour, not only the same moral words?
    \item do values still apply to the right \emph{bearers}---persons, beings, situations the bundles are about?
    \item can human observation and objection still \emph{causally change} what the system does before irreversible harm?
    \item do systems it creates or empowers inherit that structure, not just the vocabulary?
    \item can labs, markets, and institutions be steered toward regimes where correction stays possible rather than selected away?
  \end{itemize}
  \item Agents must be found operationally---without assuming in advance what the optimizer is.
  \item Human values are not arbitrary labels over situations. They appear to have compressed, hub-shaped structure: learnable value-bundle directions under a fixed ontology, with the hard failures concentrated in bearer maps, ontology shift, correction, and successor transport.
  \item A serious safety case must track \emph{correction channels}, \emph{successor constraints}, and \emph{alignment basins}.
\end{itemize}

\section*{Thesis}

Superintelligence alignment is the problem of preserving human-correctable value-bearing processes across capability growth, ontology shift, successor creation, and strategic multi-agent selection pressure, under the assumption that civilization still has enough correction capacity to participate in the process.

Load-bearing assumptions are stated in the chapters that need them and collated in Appendix~\ref{appe-assumptions}.

\section*{Rejected Simplifications}

\begin{itemize}
  \item Fixed utility functions as the sole alignment target.
  \item Alignment as a one-time training problem rather than a dynamical guarantee.
  \item Treating the AI model as the only relevant agent boundary.
\end{itemize}

Part~I develops the reframing; the Introduction states the book's five connected claims in full.

\section*{What This Book Tries to Establish}

At moderate strength: that boundary discovery, value-bundle geometry, bearer maps, correction-channel integrity, successor constraints, and alignment basins are variables that serious alignment work must make explicit---not background noise.
At stronger strength for the value-learning part: the relevant alternative to scalar reward learning is not unconstrained inference over arbitrary high-dimensional rewards, but constrained inference over compressed bundle-response geometry.
The book therefore rejects the strong pessimistic claim that useful human value structure is too arbitrary to learn at all, while preserving the weaker and more defensible claim that full human values are not safely identifiable from sparse behavior alone under ontology shift.

Progress should look like \emph{artifacts}: boundary audits, bundle evaluations, correction-channel audits, successor certification tests, adversarial measurement suites, and safety cases that list assumptions and failure modes.
The Introduction names these; later chapters and appendices instantiate them.

\section*{What This Book Does Not Claim}

This manuscript develops a conceptual and formal framework for superintelligence alignment.
It is not a claim that the alignment problem is solved.
The book distinguishes established claims, plausible hypotheses, speculative extensions, and open research problems (Appendix~\ref{apph-research-program}).

\section*{How This Book Handles External Doom Arguments}

The book does not organize itself around any one external doom argument.
Instead, it builds a positive framework and then stress-tests it against the strongest objections in a late chapter.
Yudkowsky's List of Lethalities is treated as an adversarial checklist.
Some points are answered, some are weakened by value-bundle and bearer-import structure, some are reframed as basin-transition problems, and some remain open.

\section*{Navigation}

The part-by-part map is in \textit{How to Read This Book} (Introduction).
The chapter list and draft status are in \textit{Current Status}.
The civilizational loop and argument orientation are in the Introduction; reframing detail begins in Part~I.
