\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

\section*{Why this book}

Superintelligence alignment is too large and too cross-disciplinary for a single essay or a pile of notes.
The argument runs from operational agent boundaries through value geometry, correction channels, successors, selection pressure, and safety cases---with formal dependencies between them.
A book-shaped manuscript can carry that thread without losing definitions, citations, and cross-references.

This project is primarily a \textbf{knowledge base and roadmap}: a navigable source for papers, talks, funding cases, and narrower publications extracted from a larger structure.
It does not assume conventional press publication.
Many chapters are still drafts; see \textit{Current Status} for what is stable today.

\section*{Who it is for}

The text is written for four overlapping audiences:
\begin{description}[style=nextline,leftmargin=2.5em]
\item[Alignment researchers] who need a precise framework and explicit claim strength.
\item[Safety engineers and eval builders] who need operational artifacts and audit templates.
\item[Funders and policy-adjacent readers] who need to see what decisions change if the framework is right.
\item[Capable generalists] who need a self-contained map without prior project jargon.
\end{description}
No audience is expected to accept every formalism.
Operational definitions come first; mathematics is compression, not ornament.

\section*{How to read it}

Pick an entry point:
\begin{description}[style=nextline,leftmargin=2.5em]
\item[\textbf{Executive Overview}] Two pages: thesis, rejected simplifications, what the book does not claim, and how external doom arguments are used.
\item[\textbf{Introduction}] Orientation to the argument: five connected claims, what would count as progress, the ten-part roadmap, and the practical hope.
\item[\textbf{Part~I (Chapters 1--5)}] Reframing: wrong object, civilizational loop, dynamical guarantee, fixed values, scope and assumptions (Chapter~\ref{ch:assumptions-scope-failure-coverage}).
\item[\textbf{Appendices}] Operational glossary (Appendix~\ref{appf-glossary}), assumptions index (Appendix~\ref{appe-assumptions}), notation index (Appendix~\ref{appa-notation}), research program (Appendix~\ref{apph-research-program}).
\end{description}

Load-bearing assumptions are introduced in the chapters that need them and collated in Appendix~\ref{appe-assumptions}.
The recurring decision question is: \emph{what would we audit, measure, or stop if this model is true?}

\section*{Authorship note}

Most of the current text is \textbf{largely AI-written}, produced with AI assistance under human direction, review, and editing.
Gunnar Zarncke sets thesis, scope, source canon, and revision priorities; agents draft and integrate chapter text from outlines and prior notes.
Passages should be read as structured working material until independently reviewed.
If you reuse material elsewhere, cite and attribute appropriately.
