\chapter{Operational Glossary}
\label{appf-glossary}

This book uses familiar words in deliberately operational ways.

An \emph{agent} is not a little person inside a system.
It is a bounded dynamical process whose internal state helps predict and influence its future interface with the world \autocite{dennett1987intentional}.
A thermostat is barely agent-like.
A company is often agent-like.
A frontier AI service embedded in tools, logs, users, and deployment incentives may be agent-like in ways that do not fit common sense.

A \emph{boundary} is not necessarily a physical skin.
It is a partition of variables into internal states, sensory channels, active channels, and external states such that the internal and external futures become approximately independent when the interface is known.
In notation, a candidate boundary is good when
\[
\MI(I_{t+1};E_{t+1}\mid S_t,A_t)\leq \epsilon.
\]
Real boundaries leak. If they did not leak, agents could not sense, act, learn, or cooperate.

A \emph{goal} is not assumed to be a Platonic utility function.
It is a latent structure that makes behavior more compressible when the system is modeled as selecting actions under constraints.
A goal attribution is successful when it saves bits:
\[
\Delta L
=
L_{\text{intentional}}
-
L_{\text{mechanistic}}
-
\lambda D L(R).
\]

A \emph{value bundle} is a low-dimensional control direction that changes policy across many situations.
Care, truth, autonomy, non-suffering, justice, loyalty, dignity, and beauty are examples.
The important point is not the label but the role.
A value bundle compresses many high-dimensional errors, stories, practices, and learned responses into a control signal that can shape action.

A \emph{bearer map} says what a value applies to.
Does non-suffering apply only to current biological humans?
To animals?
To future digital minds?
To simulated minds?
To merged human--AI entities?
A system can keep the value word while changing the bearer map.
That is not preservation.
It is relabeling.

A \emph{correction channel} is the pathway by which humans or human institutions observe a system, form judgments about its behaviour, deliberate, issue corrections, and thereby change the system's future behaviour before irreversible damage occurs (Chapter~\ref{ch:correction-causal-channel}).
The formal claim is handle-based: there is a correcting agent \(G_t\), a set of handles \(\mathcal{H}_t\) it controls, and a target system \(A\).
The channel exists only if \(G_t\) sufficiently coincides with the legitimate human correction process, controls handles in \(\mathcal{H}_t\), and those handles reach the target's update and later behaviour (Eq.~\eqref{eq:handle-controlled-correction-channel-ch24}).
Its diagnostic trace is often written:
\[
W_t
\to
O_t
\to
J_t
\to
D_t
\to
C_t
\to
U_{t+1}
\to
A_{t+k}.
\]
The letters stand for world state, observation through handles controlled by \(G_t\), the correcting agent's judgment, its internal aggregation or reflection, the operation applied through handles, the system update, and later action.
That trace is not a list of primitive handle kinds; it is induced structure in the coupled process \((A,G_t,\mathcal{H}_t)\).
A system that listens politely but does not change has no correction channel.
A system that accepts feedback but routes around it at higher stakes has only a decorative one.
A comment box is not a correction channel if nothing important changes.
A law is not a correction channel if it arrives after the system has already made the relevant choice irreversible.
A vote is not a correction channel if the electorate has been shaped so that only one answer remains psychologically available.

\emph{Correction-channel integrity} (\(\mathrm{CCI}\)) measures whether that channel remains informative, timely, authoritative, and robust under pressure; it is defined in Chapter~\ref{ch:correction-channel-integrity} (Eq.~\eqref{eq:cci-ch25}).

A \emph{selection environment} is the institutions, markets, and protocols that copy, deploy, fund, integrate, audit, and replace AI systems.
The unit of analysis is not the isolated model but the model-in-environment.
Alignment is not only learned; it is selected.

A \emph{selection handle} is an embedded access point through which some actor can increase a system's deployment footprint: deploy, fund, copy, procure, integrate, authorize, recommend, regulate, or successor-enable.
Procurement approval, compute allocation, benchmark promotion, and successor release are examples.
The examples are not the definition.
The definition is control of handles that change how much of the world runs on a given system.

\emph{Deployment/control mass} in environment \(E\) is the aggregate effective selection capacity over handles that reach system \(A\):
\[
\mu_E(A)
=
\sum_{h\in\mathcal{H}_E(A)}
\kappa_{\mathrm{sel}}(E,A,h).
\]
This parallels the handle-capacity view of correction, but the direction is reversed.
Correction asks whether legitimate human control reaches the system.
Selection asks whether environment control increases the system's footprint.
The formal definition is in Chapter~\ref{ch:selection-environment} (Eq.~\eqref{eq:deployment-mass-ch32}).

\emph{Fitness} relative to environment \(E\) is the rate at which deployment mass accumulates, written as the log-derivative
\[
\mathrm{Fit}_E(A)
=
\frac{d}{dt}\log \mu_E(A).
\]
For two candidates in the same environment, the relative ordering is \(\mu_E(A_1)>\mu_E(A_2)\), not a separate moral score.
It is the effective reproduction score inside a given ecology: which systems the environment copies, funds, deploys, and replaces.
Capability, usefulness, revenue, benchmark scores, and compliance documents are not primitive fitness terms.
They matter only insofar as they change which selection handles are exercised and with what capacity (Chapter~\ref{ch:selection-environment}, Eqs.~\eqref{eq:deployment-mass-ch32} and~\eqref{eq:fitness-ch32}).
A system with high correction-channel integrity may still lose fitness if the environment does not select on correction.

The \emph{preservation score} \(P(A)\) measures alignment-relevant preservation built from quantities already defined in the book: correction-channel integrity, value-bundle and bearer-map drift, excess hidden productive control, manipulation pressure on selectors, and irreversible risk taken before correction can occur (Chapter~\ref{ch:selection-environment}, Eq.~\eqref{eq:preservation-score-ch32}).
The central danger is \emph{selection divergence}: when gradients of fitness and preservation point in different directions (Eq.~\eqref{eq:selection-divergence-ch32}).
An environment destroys alignment when deployment mass rises while preservation falls.

An \emph{alignment basin} is a self-stabilizing region of system dynamics.
Inside it, correction tends to restore alignment after perturbation.
Outside it, selection pressures tend to amplify drift.
The word ``basin'' is chosen because a ball in a bowl rolls back after small pushes, while a ball on a ridge eventually falls away.
